---
title: "Setup Instructions: Load Forecasts from Local Zoltar sqlite File"
author: "Serena Wang, Matthew Cornell, Yuxin David Huang"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions
We now support loading forecast from a local zoltar sqlite file in `covidHubUtils`. To use this functionality, you need to set `source = "local_zoltar"` and provide the `local_zoltpy_path` and `zoltar_sqlite_file` parameters in `load_forecasts()`.

Please follow these steps to set up the required environment prior to using this functionality for the first time:

> Note: This feature is [currently beta](https://docs.zoltardata.com/forecasts/#download-the-zoltar-database-for-local-use-beta), so please [contact us](https://docs.zoltardata.com/#contact) to create an sqlite file for you that contains the forecasts you want to work with.  
 
1. Install Python 3.6+.
2. Install pipenv globally (i.e., without `--user`) via:
```bash
pip3 install pipenv
```

3. Set up your R environment so that the `PATH` environment variable includes both the `python3` and `pipenv` commands. You can read more about R and environment variables [here](https://db.rstudio.com/best-practices/managing-credentials/#use-environment-variables).

4. Clone the `zoltpy` package from [here](https://github.com/reichlab/zoltpy).
5. In the cloned `zoltpy` directory, run:
```bash
pipenv install
```
6. Install R libraries: `DBI` and `RSQLite` in Rstudio
7. Test installation
    - Download a `zoltar` sqlite file form [here](https://github.com/reichlab/covidHubUtils/blob/master/tests/testthat/test-data/test-load_forecasts_local_zoltar/db.sqlite3).
    - Run a test call of `covidHubUtils::load_forecasts()` with your actual absolute or relative paths. For example: 
```{r message=FALSE, eval=FALSE}
covidHubUtils::load_forecasts(
  models = "COVIDhub-ensemble", 
  locations = "US", 
  source = "local_zoltar", 
  local_zoltpy_path = "<your_zoltpy_dir>", 
  zoltar_sqlite_file = "<your_db.sqlite3_dir>")
```
